---
title: "harp news... and some features you might not know about"
author: "Andrew Singleton"
date: "MET Norway"
output: 
  revealjs::revealjs_presentation:
    theme: league
    transition: fade
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, dev.args = list(bg = "transparent"), collapse = TRUE,
  fig.height = 5
)

library(harp)
library(tidyr)
library(dplyr)
library(here)
library(forcats)
theme_presentation <- theme_set(theme_harp_midnight(base_size = 12))
theme_presentation <- theme_update(
    panel.background  = element_rect(fill = "transparent"), 
    plot.background   = element_rect(fill = "transparent"), 
    legend.background = element_rect(fill = "transparent"), 
    legend.key        = element_rect(fill = "transparent")
  )
```

<style>
  .reveal h1, .reveal h2, .reveal h3, .reveal h4, .reveal h5, .reveal h6 {text-transform:none;}
  .reveal section img { background:none; border:none; box-shadow:none; }
  .reveal table { border-width:0px; }
  .reveal table td { bordoer-width:0px; }
  .reveal pre code {background:#222;}
  .reveal pre {margin: 7px auto;}
  code span.kw {color:yellow}
  code span.dt {color: tomato}
</style>

## __New read function__
`read_forecast()`
```{r read-forecast-intro, echo = TRUE, eval = FALSE}
read_forecast(
  start_date,
  end_date,
  fcst_model,
  parameter
)
```

Replaces `read_eps_interpolate()` and `read_det_interpolate()`

...and does a whole lot more!

## __Default behaviour__
_Read data "as is"_

- For vfld, this means to just read the data
- For gridded data, the gridded fields are returned

## vfld

```{r read-forecast-vfld-code, echo=TRUE, eval = FALSE}
read_forecast(
  start_date  = 2019021700,
  end_date    = 2019021718,
  fcst_model  = "AROME_Arctic_prod",
  file_path   = system.file("vfld", package = "harpData"),
  return_data = TRUE
)
```

```{r read-forecast-vfld-res}
read_forecast(
  start_date  = 2019021700,
  end_date    = 2019021718,
  fcst_model  = "AROME_Arctic_prod",
  lead_time   = seq(0, 3),
  file_path   = system.file("vfld", package = "harpData"),
  return_data = TRUE
) %>% 
  transmute(fcdate, lead_time, SID, parameter, AA_det = AROME_Arctic_prod_det)
```

## grib
```{r read-forecast-grib-code, echo = TRUE, eval = FALSE}
read_forecast(
  2018071000,
  2018071000,
  "AROME_Arctic",
  "T2m",
  file_path     = system.file("grib/AROME_Arctic", package = "harpData"),
  file_template = "harmonie_grib_fp",
  return_data   = TRUE
)
```

```{r read-forecast-grib-res}
read_forecast(
  2018071000,
  2018071000,
  "AROME_Arctic",
  "T2m",
  lead_time     = seq(0, 3),
  file_path     = system.file("grib/AROME_Arctic", package = "harpData"),
  file_template = "harmonie_grib_fp",
  return_data   = TRUE
) %>% 
  transmute(fcdate, lead_time, parameter, AA_det = AROME_Arctic_det) %>% 
  print(n = 7)
```

## Transformations

- __interpolate__ : interpolate gridded data to points
- __regrid__ : interpolate / reproject gridded data to a new grid
- ___xsection___ _: extract a vertical cross section of the data_

## `transformation = "interpolate"`

```{r read-forecast-interpolate-code, echo = TRUE, eval = FALSE}
read_forecast(
    2018071000,
    2018071000,
    "AROME_Arctic",
    "T2m",
    transformation = "interpolate",
    file_path      = system.file("grib/AROME_Arctic", package = "harpData"),
    file_template  = "harmonie_grib_fp",
    return_data    = TRUE
)
```

```{r read-forecast-interpolate-res}
read_forecast(
    2018071000,
    2018071000,
    "AROME_Arctic",
    "T2m",
    lead_time      = seq(0, 3),
    transformation = "interpolate",
    file_path      = system.file("grib/AROME_Arctic", package = "harpData"),
    file_template  = "harmonie_grib_fp",
    return_data    = TRUE
) %>% 
  transmute(fcdate, lead_time, SID, parameter, AA_det = AROME_Arctic_det) %>% 
  print(n = 7)
```

---

### `transformation_opts = interpolate_opts()`
```{r read-forecast-interpolate-my-stations, echo = TRUE}
my_stations <- data.frame(
  SID = c(1003, 1004, 1006),
  lat = c(77.0000, 78.9167, 78.2506),
  lon = c(15.5000, 11.9331, 22.8225)
)
```

```{r read-forecast-interpolate-opts-code, eval = FALSE, echo = TRUE}
read_forecast(
  2018071000,
  2018071000,
  "AROME_Arctic",
  "T2m",
  transformation      = "interpolate",
  transformation_opts = interpolate_opts(
    stations    = my_stations,
    correct_t2m = FALSE,
    method      = "bilinear"
  ),
  file_path           = system.file("grib/AROME_Arctic", package = "harpData"),
  file_template       = "harmonie_grib_fp",
  return_data         = TRUE
)
```

---

### `transformation_opts = interpolate_opts()`
```{r read-forecast-interpolate-opts-res}
read_forecast(
  2018071000,
  2018071000,
  "AROME_Arctic",
  "T2m",
  lead_time     = seq(0, 3),
  transformation      = "interpolate",
  transformation_opts = interpolate_opts(
    stations    = my_stations,
    correct_t2m = FALSE,
    method      = "bilinear"
  ),
  file_path           = system.file("grib/AROME_Arctic", package = "harpData"),
  file_template       = "harmonie_grib_fp",
  return_data         = TRUE
) %>% 
  transmute(fcdate, lead_time, SID, parameter, AA_det = AROME_Arctic_det)
```

## `transformation = "regrid"`
```{r read-forecast-regrid-my-domain, echo = TRUE}
library(meteogrid)
my_domain <- Make.domain(
  "lambert",
  clonlat = c(15.57905, 78.21638),
  nxny    = c(15, 15), 
  dxdy    = c(10000, 10000),
  reflat  = 77.5,   
  reflon  = -25
)
```

```{r read-forecast-regrid-code, echo = TRUE, eval = FALSE}
read_forecast(
  2018071000,
  2018071000,
  "AROME_Arctic",
  "S10m",
  transformation      = "regrid",
  transformation_opts = regrid_opts(new_domain = my_domain),
  file_path           = system.file("grib/AROME_Arctic", package = "harpData"),
  file_template       = "harmonie_grib_fp",
  return_data         = TRUE
)
```

## `transformation = "regrid"`
```{r read-forecast-regrid-res}
read_forecast(
  2018071000,
  2018071000,
  "AROME_Arctic",
  "S10m",
  lead_time           = seq(0, 3),
  transformation      = "regrid",
  transformation_opts = regrid_opts(
    new_domain    = my_domain
  ),
  file_path           = system.file("grib/AROME_Arctic", package = "harpData"),
  file_template       = "harmonie_grib_fp",
  return_data         = TRUE
) %>% 
  transmute(fcdate, lead_time, parameter, AA_det = AROME_Arctic_det) %>% 
  print(n = 10)
```

## `transformation = "xsection"`
```{r read-forecast-xsection-code, echo = TRUE, eval = FALSE}
read_forecast(
  20180710, 
  20180710, 
  "AAEPS", 
  "T", 
  members             = c(0, 1), 
  vertical_coordinate = "model", 
  transformation      = "xsection", 
  transformation_opts = xsection_opts(a = c(11, 78), b = c(14, 78.25))
  file_path           = system.file("netcdf", package = "harpData"), 
  file_template       = "{fcst_model}/fc{YYYY}{MM}{DD}{HH}.nc", 
  file_format_opts    = netcdf_opts("met_norway_eps"), 
  return_data         = TRUE
)
```

## `transformation = "xsection"`
```{r read-forecast-xsection-res}
read_forecast(
  20180710, 
  20180710, 
  "AAEPS", 
  "T", 
  lead_time           = seq(0, 3),
  members             = c(0, 1), 
  vertical_coordinate = "model", 
  transformation      = "xsection", 
  transformation_opts = xsection_opts(a = c(11, 78), b = c(14, 78.25)),
  file_path           = system.file("netcdf", package = "harpData"), 
  file_template       = "{fcst_model}/fc{YYYY}{MM}{DD}{HH}.nc", 
  file_format_opts    = netcdf_opts("met_norway_eps"), 
  return_data         = TRUE
) %>% select(fcdate, lead_time, parameter, level_type, level, starts_with("AA"))
```

## Save to SQLite

_(Interpolated / point data only)_

```{r save-to-sqlite-code, eval = FALSE, echo = TRUE}
read_forecast(
  start_date       = 2019021700,
  end_date         = 2019021718,
  fcst_model       = c("MEPS_prod", "CMEPS_prod"),
  parameter        = c("T", "Td"),
  by               = 3,
  lead_time        = seq(0, 12, 3),
  members          = list(
    MEPS_prod  = seq(0, 10),
    CMEPS_prod = c(0, 1, 3, 4, 5, 6)
  ),
  lags             = list(CMEPS_prod = c(0, 0, 2, 2, 1, 1)),
  file_path        = system.file("vfld", package = "harpData"),
  file_template    = "vfld_eps",
  output_file_opts = sqlite_opts(
    path = "/path/to/sqlite/data", template = "fctable"
  )
)
```

## Vertical profile verification
```{r vertical-profile-read-fcst, echo = TRUE}
fcst <- read_point_forecast(
  start_date          = 2019021700,
  end_date            = 2019021718,
  fcst_model          = c("CMEPS_prod", "MEPS_prod"),
  fcst_type           = "EPS",
  parameter           = "T", 
  vertical_coordinate = "pressure",
  lead_time           = seq(0, 12, 3),
  by                  = "6h",
  lags                = list(
    CMEPS_prod = paste0(seq(0, 5), "h"),
    MEPS_prod  = "0h"
  ),
  file_path           = system.file("FCTABLE/ensemble", package = "harpData")
) %>% common_cases()
```

```{r vertical-profile-read-obs-join, echo = TRUE}
obs <- read_point_obs(
  first_validdate(fcst),
  last_validdate(fcst),
  "T",
  vertical_coordinate = "pressure",
  obs_path            = system.file("OBSTABLE", package = "harpData")
)

fcst <- join_to_fcst(fcst, obs)

verif <- ens_verify(fcst, T, groupings = c("leadtime", "p"))
```

---

`plot_profile_verif()`
```{r plot-profile-verif-filter, echo = TRUE}
plot_profile_verif(verif, spread_skill, filter_by = vars(leadtime == 3))
```

---

`plot_profile_verif()`
```{r plot-profile-verif-facet, echo = TRUE}
verif <- mutate_list(verif, leadtime = paste0("T + ", leadtime, "h"))
plot_profile_verif(verif, spread_skill, facet_by = vars(fct_inorder(leadtime)))
```

## Grouped verification
```{r read-forecast-group, echo = TRUE}
fcst <- read_point_forecast(
  2019021700,
  2019021718,
  "MEPS_prod",
  "EPS",
  "T2m",
  file_path = system.file("FCTABLE/ensemble", package = "harpData")
)

obs <- read_point_obs(
  first_validdate(fcst),
  last_validdate(fcst),
  "T2m",
  stations = pull_stations(fcst),
  obs_path = system.file("OBSTABLE", package = "harpData")
)
```

## Grouped verification
```{r verif-forecast-group, echo = TRUE}
obs  <- mutate(obs, region = floor(SID / 1000))

fcst <- join_to_fcst(fcst, obs)

verif <- ens_verify(
  fcst, T2m, groupings = list("leadtime", c("leadtime", "region"))
)

verif <- mutate_list(
  verif, region = fct_inorder(paste0("Region: ", region))
)
```

## Grouped verification
```{r plot-forecast-group, echo = TRUE}
plot_point_verif(verif, crps, facet_by = vars(region))
```

## Binding verifcation data
```{r prep_data_to_bind}
params = c("T2m", "S10m")
for (param in params) {
  fcst <- read_point_forecast( 2019021700,
    2019021718,
    "MEPS_prod",
    "EPS",
    param,
    file_path = system.file("FCTABLE/ensemble", package = "harpData")
  )
  
  obs <- read_point_obs(
    first_validdate(fcst),
    last_validdate(fcst),
    param,
    stations = pull_stations(fcst),
    obs_path = system.file("OBSTABLE", package = "harpData")
  )
  
  fcst <- join_to_fcst(fcst, obs)
  
  assign(paste0("verif_", param), ens_verify(fcst, {{param}}))
}
```

```{r bind_and_plot, echo = TRUE}
verif <- bind_point_verif(verif_T2m, verif_S10m)
plot_point_verif(verif, spread_skill, facet_by = vars(parameter), facet_scales = "free_y")
```

## Gridded data
```{r read-gridded-fcst, echo = TRUE}
t2m <- read_forecast(
  20180710, 
  20180710, 
  "AAEPS", 
  "T2m", 
  lead_time           = seq(0, 12, 3),
  file_path           = system.file("netcdf", package = "harpData"), 
  file_template       = "{fcst_model}/fc{YYYY}{MM}{DD}{HH}.nc", 
  file_format_opts    = netcdf_opts("met_norway_eps"), 
  return_data         = TRUE
) %>% 
  scale_point_forecast(-273.15, new_units = "degC")
```

## Gridded data
```{r print-gridded-fcst, echo = TRUE}
t2m
```

## Gridded data
```{r stats-gridded-fcst, echo = TRUE}
ens_stats(t2m)
```

## Gridded data
```{r more-stats-gridded-fcst, echo = TRUE}
ens_stats(t2m, min = TRUE, max = TRUE, prob_thresh = 0, nbh_radius = 7)
```

## Gridded data
```{r plot-stats-gridded-fcst, echo = TRUE}
t2m <- ens_stats(t2m, min = TRUE, max = TRUE, prob_thresh = 0)
plot_field(t2m, "AAEPS", ens_spread, 2018071000, 12)
```

## Gridded data
```{r plot-prob-ge0-gridded-fcst, echo = TRUE}
plot_field(t2m, "AAEPS", prob_ge_0, 2018071000, 12)
```

## Gridded data
```{r plot-prob-le0-gridded-fcst, echo = TRUE}
t2m <- ens_stats(t2m, prob_thresh = 0, prob_inequality = `<=`)
plot_field(t2m, "AAEPS", prob_le_0, 2018071000, 12, breaks = seq(0.2, 1, 0.2))
```

## Gridded data
```{r plot-prob-le0-nbh7-gridded-fcst, echo = TRUE}
t2m <- ens_stats(t2m, prob_thresh = 0, prob_inequality = `<=`, nbh_radius = 7)
plot_field(t2m, "AAEPS", prob_le_0, 2018071000, 12, breaks = seq(0.2, 1, 0.2), palette = scico::scico(10, palette = "tokyo", direction = -1))
```

## Observation errors

For ensembles, observation errors can be accounted for by including an error component in the ensemble forecast distribution. This is done with the `jitter_forecast` function. 
```{r obs-error-read-fcst}
fcst <- read_point_forecast( 
  2019021700,
  2019021718,
  "MEPS_prod",
  "EPS",
  "T2m",
  file_path = system.file("FCTABLE/ensemble", package = "harpData")
) %>% 
  join_to_fcst(
    read_point_obs(
      2019021700,
      2019022018,  
      "T2m",
      obs_path = system.file("OBSTABLE", package = "harpData")
    )
  )
```

```{r jitter-fcst, echo = TRUE}
verif        <- ens_verify(fcst, T2m)
jitter_func  <- function(x) x + rnorm(1, 0, 1)
verif_jitter <- ens_verify(jitter_fcst(fcst, jitter_func), T2m)

verif <- mutate_list(verif, `Obs error` = "No Obs errors")
verif_jitter <- mutate_list(verif_jitter, `Obs error` = "Obs errors ncluded")

verif <- bind_point_verif(verif, verif_jitter)
```

## Observation errors
```{r jitter-fcst-spread-skill, echo = TRUE}
plot_point_verif(verif, spread_skill, colour_by = `Obs error`)
```

## Observation errors
```{r jitter-fcst-rank-hist, echo = TRUE, fig.height=4}
plot_point_verif(
  verif, normalized_rank_histogram, colour_by = `Obs error`, 
  rank_is_relative = TRUE, rank_hist_type = "lollipop"
)
```

## Score cards
[See Article on harphub](https://harphub.github.io/harp/articles/scorecards.html)

## What's happening next year
- Finalize inline documentation and websites with function reference
- harp [book](https://harphub.github.io/harp_tutorial/) - Need reviewers
- Formal tests [this is a big job!]

## What's happening next year
- Minor improvements
    - Profiling of the more complicated functions for speed and memory use
    - Harmonize function arguments
    - Verification against analysis [`read_obs()` / `read_analysis()` function to compliment `read_forecast()`]
    - Simpler plotting of fields and enhanced capability (e.g. faceting)
    - Wind speed and direction from U and V
    - Verification of wind direction
    - Combined and categorical probabilities
    

## What's happening next year
- More options in shiny app
    - Plot vertical profile verification
    - Different options for top level menus
    - [Graphics better optimized for web display with popups etc.: d3.js]

- Your requests
    - File issues at the [harphub repository](https://github.com/harphub)
    - [harpIO](https://github.com/harphub/harpIO/issues) - for IO, data structures and data wrangling
    - [harpPoint](https://github.com/harphub/harpPoint/issues) - for point verification
    - [harpVis](https://github.com/harphub/harpVis/issues) - for visualization and shiny apps
    - [harpSpatial](https://github.com/harphub/harpSpatila/issues) - for spatial verification
    - [harp](https://github.com/harphub/harp/issues) - for issues that cover more than 1 harp package
