---
title: "harp version 0.3"
author: "Andrew Singleton"
format: 
  revealjs:
    theme: [default, custom.scss]
    embed-resources: true
    code-line-numbers: false
---

# General improvements

-   Major overhaul (refactor / rewrite) of verification functions
-   Verification for quantile thresholds
-   More memory efficient, possibly faster
-   More feedback on what's happening
-   Can compute new scores without having to edit internal code
-   New "do (nearly) everything" functionalities

# Running a verification job

## In an R session

-   `run_point_verif()`

```{r eval=FALSE, echo=TRUE}
run_point_verif(
  dttm       = seq_dttm(20251010, 20251020, "3h"),
  fcst_model = c("MEPS", "IFSENS"),
  param_defs = make_verif_param("S10m", verif_thresholds = seq(2.5, 10, 2.5)),
  fcst_path  = "/data/harp/FCTABLE",
  obs_path   = "/data/harp/OBSTABLE",
  out_path   = "/data/harp/verification"
)
```

## Helper functions

-   `make_verif_param()`
    -   param name
    -   param names in files
    -   scaling: `make_scaling()`
    -   gross error check bounds
-   `make_verif_defaults()`
    -   groupings: `make_verif_groups()`
    -   error check against obs sd

## Batch running

-   `new_point_verif_project()`
-   DEMO!

# New ensemble scores

## Threshold weighted CRPS

-   Computes CRPS emphasizing cases the fall within the threshold

-   "Clamps" data outside the threshold of interest

## Threshold weighted CRPS

```{r, fig.width=12, fig.align='center'}
library(dplyr)
library(ggplot2)

dd <- data.frame(
  fc = sort(rnorm(2.5e4)),
  ob = 0,
  y = seq_len(2.5e4),
  th = 0.25,
  type = "crps",
  th_min = -0.5, 
  th_max = 0.5
)

dd$fc_btw <- dd$fc

dd <- rbind(
  dd, 
  dplyr::mutate(
    dd, 
    fc = dplyr::case_when(fc >= th ~ fc, .default = th),
    fc_btw = dplyr::case_when(
      fc_btw >= th_max ~ th_max, fc_btw <= th_min ~ th_min, .default = fc_btw
    ),
    type = "tw_crps"
  )
)

ggplot(dd, aes(y = y)) + 
  geom_ribbon(aes(xmin = ob, xmax = fc), fill = "#FFBB55") +
  geom_line(aes(x = fc), colour = "#FF6600", linewidth = 3) + 
  geom_line(aes(x = ob), linewidth = 2, colour = "white") + 
  geom_line(aes(x = th), linewidth = 1.5, linetype = 2, colour = "white") + 
  facet_wrap(~toupper(type), ncol = 2) + 
  coord_cartesian(xlim = c(-2.5, 2.5)) + 
  harpVis::theme_harp_map() + 
  theme(
    plot.background = element_rect(fill = "#333333", colour = "transparent"), 
    panel.background = element_rect(fill = "#333333"),
    strip.background = element_rect(fill = "#333333"),
    strip.text = element_text(colour = "#FF6600", size = 18)
  )

```

## Threshold weighted CRPS

```{r, fig.width=12, fig.align='center'}
ggplot(dd, aes(y = y)) + 
  geom_ribbon(aes(xmin = ob, xmax = fc_btw), fill = "#FFBB55") +
  geom_line(aes(x = fc_btw), colour = "#FF6600", linewidth = 3) + 
  geom_line(aes(x = ob), linewidth = 2, colour = "white") + 
  geom_line(aes(x = th_max), linewidth = 1.5, linetype = 2, colour = "white") + 
  geom_line(aes(x = th_min), linewidth = 1.5, linetype = 2, colour = "white") + 
  facet_wrap(~toupper(type), ncol = 2) + 
  coord_cartesian(xlim = c(-2.5, 2.5)) + 
  harpVis::theme_harp_map() + 
  theme(
    plot.background = element_rect(fill = "#333333", colour = "transparent"), 
    panel.background = element_rect(fill = "#333333"),
    strip.background = element_rect(fill = "#333333"),
    strip.text = element_text(colour = "#FF6600", size = 18)
  )


```

## UUI spread skill

$$\mathbb{E}[STDE_{P}] = 2\mathbb{E}[VARE_{P}]$$

## UUI spread skill

$$\mathbb{E}[STDE_{P}] = 2\mathbb{E}[VARE_{P}]$$

-   Based on the notion that the unbiased measure of spread is the expected difference between any two ensemble members $$
    \small
    \begin{aligned}
    \mathbb{E}[(Y_{1}-Y_{2})^{2}]
    &=\mathbb{E}[Y_{1}-\mu+\mu-Y_{2}^{2}]\\
    &=\mathbb{E}[(Y_{1}-\mu)^{2}]-2\mathbb{E}[Y_{1}-\mu]\mathbb{E}[Y_{2}-\mu]+\mathbb{E}[(Y_{2}-\mu)^{2}]\\
    &=2\mathbb{E}[(Y-\mu)^{2}]-0\\
    &=2Var(Y)
    \end{aligned}
    $$ The 3rd equality comes from the fact $Y_{1}$ and $Y_{2}$ have the same distribution as $Y$

# Plotting fields

## New plot method

-   `plot(data, col = )`
-   Uses *ggplot* so easily modifiable
-   Demo!

## Wind vectors

-   `geom_geowindvec(aes(u = , v = ), skip = )`
-   Skipping is important!
-   Demo!

## Neighbourhood probabilities

-   `ens_prob()` for pixel probabilities
-   `nbhd_smooth()` for neighbourhood probabilities
-   Needs to be done member by member and aggregated
-   Will (hopefully) be built into `ens_prob()` in the future
-   Need to add `ens_pfss()` function
-   Demo!

# Cross Sections

## Cross sections

-   Use `transformation = "xsection"` + `xsection_opts()`
-   `geom_xs()` : model levels
-   `geom_xs_pressure()` : model levels -\> vertical pressure
    -   needs:
        -   surface pressure

## Cross sections

-   `geom_xs_height()` : model levels -\> height ASL
    -   needs:
        -   surface pressure
        -   temperature on model levels
        -   specific humidity on model levels
        -   representation of topographic height (usually surface geopotential)
-   Demo!

# Future plans

## 

-   Just maintenance 😞
