---
title: "An introduction to harp"
author: "Andrew Singleton"
date: "SUV verifikasjonsmøte 19 March 2021"
output: 
  revealjs::revealjs_presentation:
    theme: league
    transition: fade
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, dev.args = list(bg = "transparent"), collapse = TRUE,
  fig.height = 5
)

library(harp)
library(tidyr)
library(dplyr)
library(here)
library(forcats)

theme_presentation <- theme_set(theme_harp_midnight(base_size = 12))
theme_presentation <- theme_update(
    panel.background  = element_rect(fill = "transparent"), 
    plot.background   = element_rect(fill = "transparent"), 
    legend.background = element_rect(fill = "transparent"), 
    legend.key        = element_rect(fill = "transparent")
  )
```

<style>
  .reveal {font-family: comfortaa;}
  .reveal h1, .reveal h2, .reveal h3, .reveal h4, .reveal h5, .reveal h6 {text-transform:none; font-family:comfortaa;}
  .reveal section img { background:none; border:none; box-shadow:none; }
  .reveal table { border-width:0px; }
  .reveal table td { bordoer-width:0px; }
  .reveal pre code {background:#222;}
  .reveal pre {margin: 7px auto;}
  code span.kw {color:yellow}
  code span.dt {color: tomato}
</style>

## __What is harp?__

- A "framework" for working with meteorological / climatological data in R
- Comprises several packages:
  - harpIO - for reading / writing data
  - harpPoint - for point verification
  - harpSpatial - for spatial verification
  - harpVis - for visualization
  - (harpMethods) - for underlying classes, methods and common functionality
  - harp - meta package for functionality that includes more than 1 harp pacakge
  - harpData - for example data
  
## __Philisophy__

- harp is built with "tidy" data principles in mind
- No matter where the data come from or what they represent the structure is (nearly) always the same
  - Data are stored as data frames
  - Every column is a "variable"
  - Every row is an "observation"
  - Every cell is a "value"
- Designed to integrate with the "tidyverse" in R
  - data wrangling with _dplyr_ and _tidyr_ verbs
  - plotting with _ggplot2_ (where appropriate)
  - functional programming with _purrr_

## __Installation__

- The harp code is stored on Github under "harphub"
- Can be installed in R with the following:

```{r install, eval=FALSE, echo=TRUE}
install.packages("remotes")
library(remotes)
install_github("harphub/harp")
```

- Installation can take some time as harp has many (too many?) dependencies

## __Reading forecast data__

`read_forecast()` can be used to read forecasts from any file format.

```{r read-forecast1, echo=TRUE, eval=FALSE}
library(harp)

read_forecast(
  start_date,
  end_date,
  fcst_model,
  (parameter),
  (file_path),
  (file_template),
  (file_format_opts),
  (lead_time),
  (members)
)
```

## File name templates

- File names for `read_forecast()` are generated from templates. Dynamic parts of the file name are in { }

```{r file-name-template, echo=TRUE, eval=FALSE}
file_path  <- "/my/path"
fcst_model <- "MEPS"
start_date <- 2021031900 
end_date   <- 2021031912
by         <- "6h"
lead_time  <- seq(0, 3, 3)
members    <- c(0, 2)
template   <- "{fcst_model}/{YYYY}/{MM}/{DD}/{fcst_model}_fc{YYYY}{MM}{DD}T{HH}Z_mbr{MBR3}+{LDT3}"
```

## File name templates

- File names for `read_forecast()` are generated from templates. Dynamic parts of the file name are in { }

```{r file-name-generate, echo=TRUE, eval=FALSE}
file_path  <- "/my/path"
fcst_model <- "MEPS"
start_date <- 2021031900 
end_date   <- 2021031912
by         <- "6h"
lead_time  <- seq(0, 3, 3)
members    <- c(0, 2)
template   <- "{fcst_model}/{YYYY}/{MM}/{DD}/{fcst_model}_fc{YYYY}{MM}{DD}T{HH}Z_mbr{MBR3}+{LDT3}"
```

```{r file-name-template-out}
file_path  <- "/my/path"
fcst_model <- "MEPS"
start_date <- 2021031900 
end_date   <- 2021031906
by         <- "6h"
lead_time  <- seq(0, 3, 3)
members    <- c(0, 2)
template   <- "{fcst_model}/{YYYY}/{MM}/{DD}/{fcst_model}_fc{YYYY}{MM}{DD}T{HH}Z_mbr{MBR3}+{LDT3}"
generate_filenames(
  file_path     = file_path,
  file_template = template,
  fcst_model    = fcst_model,
  start_date    = start_date,
  end_date      = end_date,
  lead_time     = lead_time,
  members       = members,
  by            = by
) %>% cat(sep = "\n")
```

## Built in templates

- For "known" file name structures there are built in templates

```{r show-file-templates, echo=TRUE, eval=FALSE}
show_file_templates()
```

```{r show-file-templates-output}
show_file_templates()
```

## Built in templates

- To show a full template, pass the row from the output to `show_file_templates()`.

```{r show-file-templates-meps, echo=TRUE}
show_file_templates(21)
```

## The `harp_fcst` object

- The output of `read_forecast()` is a `harp_fcst` object - named list of data frames - one for each `fcst_model`

```{r read-meps, eval=FALSE, echo=TRUE}
fcst <- read_forecast(
  2021031200,
  2021031212,
  "MEPS",
  "Pcp",
  lead_time        = seq(12, 24, 12),
  members          = c(0, 10, 20),
  file_template    = "meps_lagged_6h_subset",
  file_format_opts = netcdf_opts("met_norway_eps"),
  show_progress    = TRUE, 
  return_data      = TRUE
)
```

## The `harp_fcst` object

```{r show-harp-fcst}
fcst <- readRDS("fcst_data.RDS")
fcst
```

## `dplyr` verbs on `harp_fcst` objects

```{r harp-fcst-selected-cols, echo=TRUE}
fcst %>% 
  select(fcdate, lead_time, starts_with("MEPS"))
```

```{r harp-fcst-filter, echo = TRUE}
fcst %>% 
  select(fcdate, lead_time, MEPS_mbr000) %>% 
  filter(lead_time == 12)
```

## Plotting a field

```{r plot_field, echo = TRUE}
plot_field(
  fcst, fcst_model = "MEPS", plot_col = MEPS_mbr000, 
  fcdate = 2021031200, lead_time = 12
)

```

## Plotting a field

```{r plot_field-better-breaks, echo = TRUE}
plot_field(
  fcst, fcst_model = "MEPS", plot_col = MEPS_mbr000, 
  fcdate = 2021031200, lead_time = 12,
  palette = c("transparent", viridis::viridis(255)),
  breaks = c(0, 0.125, 0.25, 0.5, 1, 2, 4, 8, 16, 32, 64)
)
```

## Transformations

- `read_forecast()` can transform the data by interpolating to points, to a new grid (and / or projection), or to a vertical cross section (soonish!)

```{r transform-interpolate, echo=TRUE, eval=FALSE}
fcst <- read_forecast(
  2021031200,
  2021031212,
  c("MEPS", "AROME_ARCTIC"),
  "T2m",
  lead_time        = seq(0, 12, 3),
  file_template    = list(
    MEPS         = "meps_det", 
    AROME_ARCTIC = "arome_arctic_extracted"
  ),
  file_format_opts = netcdf_opts("met_norway_det"),
  transformation   = "interpolate", 
  show_progress    = TRUE, 
  return_data      = TRUE
)
```

## Interpolated data

```{r transform-interpolate-get-data}
fcst <- readRDS("fcst_point_data.RDS")
```

```{r, print-interpolated-data, echo=TRUE}
fcst %>% 
  select(fcdate, lead_time, SID, ends_with("_det"))
```

## Saving interpolated data

- It takes a long time to read and interpolate full model fields, so the interpolated data can be saved to SQLite files
- Simply add `output_file_opts = sqlite_opts(path = <path>)`

```{r, save-to-sqlite, echo=TRUE, eval=FALSE}
read_forecast(
  2021031200,
  2021031212,
  c("MEPS", "AROME_ARCTIC"),
  "T2m",
  lead_time        = seq(0, 12, 3),
  file_template    = list(
    MEPS         = "meps_det", 
    AROME_ARCTIC = "arome_arctic_extracted"
  ),
  file_format_opts = netcdf_opts("met_norway_det"),
  transformation   = "interpolate", 
  output_file_opts = sqlite_opts(path = "/path/to/output"),
  show_progress    = TRUE
)
```

## Reading the interpolated forecasts

- The function `read_point_forecast()` is used to read interpolated data

```{r read-point-forecast-not-run, echo=TRUE, eval=FALSE}
fcst <- read_point_forecast(
  2021031200, 
  2021031212, 
  fcst_model = c("MEPS", "AROME_ARCTIC"), 
  fcst_type  = "det", 
  parameter  = "T2m", 
  file_path  = "suv_verifikasjonsmøte_202103"
)
```

```{r read-point-forecast-is-run, echo=FALSE}
fcst <- read_point_forecast(
  2021031200, 
  2021031212, 
  fcst_model = c("MEPS", "AROME_ARCTIC"), 
  fcst_type  = "det", 
  parameter  = "T2m", 
  file_path  = getwd()
)
```

## Reading the interpolated forecasts

```{r show-interpolated-fcst, echo=TRUE}
fcst %>% 
  select(fcdate, leadtime, SID, ends_with("det"))
```

## Observations

- harp is designed to work with `vobs` data 
  - text files used with "monitor" for operational verification in HIRLAM institutes
  - available on:
    - ecgb at `/scratch/ms/dk/nhz/oprint/OBS4`
    - lustre for AROME-ARCTIC
    - Extreme security metcoop servers for MEPS (Also in SQLite)
- Interface with `frost` coming soon(ish) [based on miIO]
- Convert `vobs` to `SQLite` with `read_obs_convert()`
- Read observations from `SQLite` with `read_point_obs()`

## Point verification workflow

- Read forecasts
- Read observations
- Join
- Verify
- Visualize

## Point verification workflow

- Read forecasts : `read_point_forecast()` 
- Read observations 
- Join  
- Verify
- Visualize

## Point verification workflow

- Read forecasts : `read_point_forecast()`
- Read observations : `read_point_obs()`
- Join
- Verify
- Visualize

## Point verification workflow

- Read forecasts : `read_point_forecast()`
- Read observations : `read_point_obs()`
- Join : `join_to_forecast()`
- Verify
- Visualize

## Point verification workflow

- Read forecasts : `read_point_forecast()`
- Read observations : `read_point_obs()`
- Join : `join_to_forecast()`
- Verify : `det_verify()` / `ens_verify()`
- Visualize

## Point verification workflow

- Read forecasts : `read_point_forecast()`
- Read observations : `read_point_obs()`
- Join : `join_to_forecast()`
- Verify : `det_verify()` / `ens_verify()`
- Visualize : `plot_point_verif()`

## Point verification workflow

- Read forecasts : `read_point_forecast()`
- Read observations : `read_point_obs()`
- Join : `join_to_forecast()`
- Verify : `det_verify()` / `ens_verify()`
- Visualize : `plot_point_verif()`
- Save : `save_point_verif()`
- Explore : `shiny_plot_point_verif()`

## [Demonstration](https://andrew-met.github.io/harp_presentations/verif_demo.html)

## Resources

This presentation


<font size=5>https://andrew-met.github.io/harp_presentations/suv_verif_202103.html</font>


- [harp website](https://harphub.github.io/harp)
- [harp tutorial](https://harphub.github.io/harp_tutorial)
- [R for Data Science](https://r4ds.had.co.nz/)
