---
title: "harp's new functions for production"
author: "Andrew Singleton"
format: 
  revealjs:
    theme: [default, custom.scss]
    embed-resources: true
    code-line-numbers: false
---

# One function to rule them all

## Main verification function

`run_point_verif()` runs a point verification task for one or more parameters. 
It reads forecasts and observations from files and computes verification scores. 
Deterministic or ensemble scores are computed depending on the data.

## Helper functions

- Default groupings and error checking are specified using the `defaults` 
argument with the help of `make_verif_defualts()`. 

- Parameter names, thresholds, comparators, and scaling as well as parameter 
specific error checking and groupings are specified with the `params` argument 
with the the help of `make_verif_param()`. 

# Defaults

## Using helper functions

```{r}
library(harp)
```

```{r, eval=FALSE, echo=TRUE}
make_verif_defaults(
  verif_groups = ...,
  obs_error_sd = ...
)
```

There's also a helper function for defining groupings
  
```{r, eval=FALSE, echo=TRUE}
make_verif_groups(
  time_groups = ...,
  groups      = ...
)
```

## Example defaults

```{r, echo=TRUE}
make_verif_defaults(
  verif_groups = make_verif_groups(
    time_groups = c("lead_time", "valid_hour"), 
    groups      = c("fcst_cycle", "stn_group")
  ),
  obs_error_sd = 6
)
```

# Parameter definitions

## 

```{r, echo=TRUE, eval=FALSE}
make_verif_param(
  param,
  fcst_param          = NULL,
  fcst_scaling        = NULL,
  fcst_jitter_func    = NULL,
  obs_param           = NULL,
  obs_scaling         = NULL,
  obs_min             = NULL,
  obs_max             = NULL,
  obs_error_sd        = NULL,
  verif_groups        = NULL,
  verif_thresholds    = NULL,
  verif_comparator    = "ge",
  verif_comp_inc_low  = TRUE,
  verif_comp_inc_high = TRUE,
  verif_circle        = NULL,
  verif_members       = TRUE,
  vertical_coordinate = NA_character_
)
```

## Example parameter

```{r, echo=TRUE, eval=FALSE}
make_verif_param(
    "2m Temperature",
    fcst_param          = "T2m",
    fcst_scaling        = make_scaling(-273.15, "degC"), 
    obs_param           = "T2m",
    obs_scaling         = make_scaling(-273.15, "degC"),
    obs_min             = 223, 
    obs_max             = 333, 
    obs_error_sd        = 6, 
    verif_thresholds    = list(
      c(-30, seq(-20, 30, 5)), 
      c(-Inf, -30, seq(-20, 30, 5), Inf)
    ),
    verif_comparator    = c("ge", "between"), 
    verif_comp_inc_low  = TRUE,
    verif_comp_inc_high = FALSE
  )
```

## Example parameter

```{r}
make_verif_param(
    "2m Temperature",
    fcst_param          = "T2m",
    fcst_scaling        = make_scaling(-273.15, "degC"), 
    obs_param           = "T2m",
    obs_scaling         = make_scaling(-273.15, "degC"),
    obs_min             = 223, 
    obs_max             = 333, 
    obs_error_sd        = 6, 
    verif_thresholds    = list(
      c(-30, seq(-20, 30, 5)), 
      c(-Inf, -30, seq(-20, 30, 5), Inf)
    ),
    verif_comparator    = c("ge", "between"), 
    verif_comp_inc_low  = TRUE,
    verif_comp_inc_high = FALSE
  )
```

## `harp_verif_param`s can be concatenated

```{r, echo=TRUE, eval=FALSE}
params <- c(
  make_verif_param("2m Temperature", ...),
  make_verif_param("10m Wind Speed", ...)
)
```

# Running a verification

## The function

```{r, echo=TRUE, eval=FALSE}
run_point_verif(
  dttm,
  fcst_model,
  params,
  lead_time            = seq(0, 48, 3),
  members              = NULL,
  lags                 = "0s",
  stations             = NULL,
  station_groups       = NULL,
  main_fcst_model      = NULL,
  ens_mean_as_det      = FALSE,
  dttm_rounding        = NULL,
  dttm_rounding_dirn   = c("nearest", "up", "down"),
  dttm_rounding_offset = 0,
  fcst_path            = getwd(),
  fcst_template        = "fctable",
  fcst_format          = "fctable",
  obs_path             = getwd(),
  obs_template         = "obstable",
  obs_format           = "obstable",
  out_path             = NULL,
  out_template         = "point_verif",
  out_format           = "rds",
  defaults             = make_verif_defaults(),
  return_data          = TRUE
)
```

## Main forecast model

- To speed up data reading a "main" model can be specified
- Only station IDs that exist for this model will be read for other forecast models
- An initial step identifying common cases is still done to minimize memory usage

## Rounding date-times

- For grouping by `valid_dttm`
- You may not want every time point - output too noisy
- `dttm_rounding` rounds all `valid_dttm` to nearest multiple
  - Could be e.g. `"1d"`, `"12h"`, `"6h"` etc.

## Rounding date-times

- `dttm_rounding_dirn` controls how the times are aggregated
  - `"nearest"` is centred on `dttm_rounding`
  - `"up"` is always times in the window up to and including `dttm_rounding`
  - `"down"` is always values in the window starting at `dttm_rounding`

## Rounding date-times

- `dttm_rounding_offset` offsets the value of `dttm_rounding`. 
  - If `dttm_rounding = "1d"` the aggregation will be centred on 00 UTC.
  - With `dttm_rounding_offset = "12h"` the aggregation will be centred on 12 UTC
  
# Configuration and Running

## Configuration file

- Currently there is no configuration file
- Should it be YAML à la ACCORD verification scripts?
- What variables should be included

## Running from shell

- Experimental script to be run with RScript
- Shell script to call R script
- Script will always take date-times and model names
- Other options will override what is specified in configuration file

# Availability

## harp repository

- The new functions are part of the __harp__ package
- Combine functions from _harpCore_, _harpIO_ and _harpPoint_
- Currently in _develop_ branch
- `verify_spatial()` to be moved to __harp__ as `run_spatial_verif()`?
  - Better separation of concerns for packages
  
